Staged Abstract Interpreters

### 2019-09-11 paper/50_stageabs.tex
polish

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter. Now we begin describing the implementation of
their confluence -- a staged abstract interpreter.  Unsurprisingly, the staged
abstract interpreter in this section uses the same abstract semantics from the
unstaged version in Section~\ref{unstaged_abs}.  The approach we use to refactor
the unstaged one to the staged abstract interpreter is modular, and does not
sacrifice soundness or precision. The designer of the analyzer therefore has no
need to rewrite the analysis. We first present the staged lattices and staged
monads, and then discuss the staged version of primitive operations, especially
@close@, @ap_clo@ and @fix@. In the end, we discuss several optimizations.

@@

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve the stage polymorphism. Now we instantiate the type @R@ as @Rep@ and
still use the power-set as example to describe its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

The trait @RepLattice@ is shown by instantiating @Lattice@ with the type @Rep@.
The @RepSetLattice@ for power sets is an instance of @RepLattice@, where the
lattice operations are eventually delegated to the operations on the staged set
data type, such as @subsetOf@, @union@ and @intersect@. We do not need to
change the implementations, except the types -- such as changing them from
@Set[T]@ to @Rep[Set[T]]@.  We show the changes in light gray code. 
%The LMS library provides implicit conversions that lift the current-stage values to
%their next-stage representations. 
In the code generation part, these operations on staged sets emit their
corresponding next-stage code. Again, other lattice structures such as products
and maps can be implemented in a similar way.

@@

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers with their staged versions. The
following code snippet shows this change.
We manually fuse the three inner transformers (@IdM@ is omitted) into a single
monad @RepSetReaderStateM[R, S, A]@, where @R@ is the type parameter for the
reader effects, and @S@ is the type for the state.  In our monad stack scheme,
@R@ and @S@ are both instantiated with the type @Cache@. Similar to the unstaged
version, the grounded result type is a pair of two staged values:
@Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ... }
\end{lstlisting}

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that where the evaluation @ev(e)@ is invoked are shifted from @ap_clo@ to
@close@, where @e@ is the body expression of a $\lambda$-term.
In other words, when staging, we eagerly specialize the interpreter and
generate code every time when we reach a $\lambda$-term, instead of lazily
calling @ev@ when the applications happen.  Similarly, here we use the same way
to handle staged $\lambda$-terms.  Additionally, in the staged abstract
interpreter, we have to handle nondeterminism incurred by the
over-approximation of runtime behavior.  In the rest of this part, we keep our
roadmap and focus on discussing closure representations and function
applications for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside @f@, we collapse the monadic value of
type @Ans@ to grounded values of type @Result@, by providing the desired
arguments, i.e., the new environment, new store, and two caches. The collapsing
of monadic values happens at the current-stage, so the invocation of @ev@ is unfolded
within staging. Finally, we generate a singleton set containing the compiled
next-stage closure (@emit_compiled_clo@), which is represented by an IR node
in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ)) }
\end{lstlisting}

@@

\paragraph{Staged Caching and Fixpoint Iteration} The fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @in/out@ cache contains the current
configuration will produce next-stage Boolean values, i.e., of type
@Rep[Boolean]@, and the branching condition cannot be determined statically.
We have to generate code for the whole @if@ expression.
Figure \ref{fig:staged_coind_cache} shows the staged version of @fix@. The
variable @res@ represents the next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of the query result
form the @out@ cache, and the @out@ cache itself. The else branch constructs a
monadic value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@
cache.  After which, we use a similar technique that collapses the monadic
value @m@ to grounded values, by providing its desired environment and etc.
Finally, we have a current-stage representation of the future-stage values
@res@, and we reify the content of @res@ back into the current-stage monad stack.

@@

Our staging schema works by unfolding the interpreter over the abstract syntax
tree of the input program. In practice, however, it would suffer from code
explosion when analyzing (specializing) large programs, which increases the
compile time. If we generate next-stage programs running on JVM, such large
generated programs would also potentially cause runtime GC overhead. In this
section, we present optimizations that largely mitigate these issues. For all
that, implementing these optimizations do not need to change the generic
interpreter.

@@

In the staged interpreters, all instances of @Env@ and @Store@ are staged. The
data structures representing these components are treated as black-boxes, i.e.,
@Rep[Map]@, which means that the operations on a @Rep[Map]@ directly become
nex-stage code, and we do not inspect any further inside. As we identified when
introducing the generic interface, the keys of an @Env@ are string-represented
identifiers in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures for environments.
For example, let us assume that the @Map[K, V]@ is implemented as a hash-map.
If all the keys of type @K@ are all known statically, then the indices for
those keys can also be computed statically. Thus, the specialized map would be
an array of type @Array[Rep[V]]@, whose elements are next-stage values, and the
size of the array is known statically as the program has finite number of
identifiers. As result, an access to the map is staged into an access to the array
with the pre-computed index.

>>>>

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter. Now we begin describing the implementation of
their confluence -- a staged abstract interpreter.  Unsurprisingly, the staged
abstract interpreter in this section uses the same abstract semantics as the
unstaged version in Section~\ref{unstaged_abs}.  The approach we use to refactor
the unstaged one to the staged abstract interpreter is modular, and does not
sacrifice soundness or precision. The designer of the analyzer therefore does not
have to rewrite the analysis. We first present the staged lattices and staged
monads, and then discuss the staged version of primitive operations, especially
@close@, @ap_clo@ and @fix@. In the end, we discuss several optimizations.

@@

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve stage polymorphism. Now we instantiate the type @R@ as @Rep@ and
still use the power-set as example to describe its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

The trait @RepLattice@ is shown by instantiating @Lattice@ with the type @Rep@.
For all type @T@, method @RepSetLattice@ provides an instance of @RepLattice@
for @Set[T]@, where the lattice operations are eventually delegated to the
operations on the staged set data type, such as @subsetOf@, @union@ and
@intersect@. 
Compared with the unstaged version, we do not need to change the
implementations, except the types -- such as changing them from @Set[T]@ to
@Rep[Set[T]]@.  We show the changes in light gray code. 
%The LMS library provides implicit conversions that lift the current-stage values to
%their next-stage representations. 
In the code generation part, these operations on staged sets emit their
corresponding next-stage code. Again, other lattice structures such as products
and maps can be implemented in a similar way.

@@

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure as the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers with their staged versions. The
following code snippet shows this change.
We manually fuse the three inner transformers (@IdM@ is omitted) into a single
monad @RepSetReaderStateM[R, S, A]@, where @R@ is the type parameter for the
reader effects, and @S@ is the type for the state.  In our monad stack scheme,
@R@ and @S@ are both instantiated with the type @Cache@. Similar to the unstaged
version, the grounded result type is a pair of two staged values:
@Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ... }
\end{lstlisting}

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that where the evaluation @ev(e)@ is invoked are shifted from @ap_clo@ to
@close@, where @e@ is the body expression of a $\lambda$-term.
In other words, when staging, we eagerly specialize the interpreter and
generate code every time we reach a $\lambda$-term, instead of lazily
calling @ev@ when the applications happen.  Similarly, here we use the same way
to handle staged $\lambda$-terms.  Additionally, in the staged abstract
interpreter, we have to handle nondeterminism incurred by the
over-approximation of runtime behavior.  In the rest of this part, we keep our
roadmap and focus on discussing the closure representation and function
application for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function
@f@, which takes four next-stage values, including the argument and latest
store, and @in/out@ caches additionally. Inside @f@, we collapse the monadic
value of type @Ans@ to grounded values of type @Result@, by providing the
desired arguments, i.e., the new environment, new store, and two caches. The
collapsing of monadic values happens at the current-stage, so the invocation of
@ev@ is unfolded during staging. Finally, we generate a singleton set
containing the compiled next-stage closure (@emit_compiled_clo@), which is
represented by an IR node in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ)) }
\end{lstlisting}

@@

\paragraph{Staged Caching and Fixpoint Iteration} The fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @in/out@ cache contains the current
configuration will produce next-stage Boolean values, i.e., of type
@Rep[Boolean]@, and the branching condition cannot be determined statically.
We have to generate code for the whole @if@ expression.
Figure \ref{fig:staged_coind_cache} shows the staged version of @fix@. The
variable @res@ represents the next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of the query result
form the @out@ cache, and the @out@ cache itself. The else branch constructs a
monadic value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@
cache.  After, we use a similar technique that collapses the monadic
value @m@ to grounded values, by providing its desired environment and etc.
Finally, we have a current-stage representation of the future-stage values
@res@, and we reify the content of @res@ back into the current-stage monad stack.

@@

Our staging schema works by unfolding the interpreter over the abstract syntax
tree of the input program. In practice, however, it would suffer from code
explosion when analyzing (specializing) large programs, which increases
compile time. If we generate next-stage programs running on the JVM, such large
generated programs would also incur GC overhead at runtime. In this
section, we present optimizations that largely mitigate these issues. 
Implementing all of those optimizations do not need to change the
generic interpreter.

@@

In the staged interpreters, all instances of @Env@ and @Store@ are staged. The
data structures representing these components are treated as black-boxes, i.e.,
@Rep[Map]@, which means that the operations on a @Rep[Map]@ directly become
next-stage code, and we do not inspect any further inside. As we identified when
introducing the generic interface, the keys of an @Env@ are string-represented
identifiers in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures for environments.
For example, let us assume that the @Map[K, V]@ is implemented as a hash-map.
If all the keys of type @K@ are all known statically, then the indices for
those keys can also be computed statically. Thus, the specialized map would be
an array of type @Array[Rep[V]]@, whose elements are next-stage values, and the
size of the array is known statically as the program has finite number of
identifiers. As result, an access to the map is staged into an access to the array
with the pre-computed index.
